Forcing for Isabelle

### 2018-02-13 blueprint.tex
atomic formulas seem complete

\renewcommand{\PP}{\mathbb{P}}
\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
\newcommand{\tyi}{\mathtt{i}}
\newcommand{\tyo}{\mathtt{o}}
\newcommand{\forceisa}{\mathtt{forces}}
\DeclareMathOperator{\val}{val}

@@

\subsection{Forcing for atomic formulas}
The base case of the definition of the forcing relation, that is, for
atomic formulas, is (in)famously the most complicated one. 
The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
following, with two mutually recursive clauses:
\begin{multline}
  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}
  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}
Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
to have a single, more complicated, clause for $t_1=t_2$ using
well-founded recursion on $t_1$
and $t_2$ and then define the case $t_1\in t_2$ explicitly.

In any case, these definitions depend on the theorem of well-founded
recursion,
so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
must be obtained from the proof of that theorem. Below, a defined
relation $R$ is \emph{set-like} if $R^{-1}\{a\}$ is a set for every $a$.
%
\begin{theorem}[{\cite[I.9.11]{kunen2011set}}]
  Assume that $R$ is a
  set-like well-founded defined relation on a class $A$, assume that 
  $\forall  x,s\,\exists!y\,\phi(x,s,y)$ and write $G(x,s)$ for the
  only $y$ such that $\phi(x,s,y)$. Then there exists another formula 
   $\psi$ such that $\forall x\,\exists!y\,\psi(x,y)$ and if we write
  $F(x)$ for the only $y$ satisfying $\psi(x,y)$, we have
  \[
  \forall a: F(a) = G(a,F\restr R^{-1}\{a\}).
  \]
\end{theorem}
%
Observe that since $F$ is a definable function, its restriction to the
set $R^{-1}\{a\}$ must also be a set by the Replacement Axiom.

There are several steps to apply this theorem to justify the
definition of forcing for atomic formulas:
\begin{enumerate}
\item As mentioned before, $F$ will be the characteristic function of
  the predicate $\forceisa$.
  %
\item The class $A$ is $V\times V = \{\lb x,y\rb : x,y\in V\}$,
  i.e.\ the class of all ordered pairs. The relation $R$ must be
  defined.
  %
\item Write down explicitly $\psi$ as an element of
  $\formula$. 
\end{enumerate}
%
Luckily, the definition of $R$ and the formula $\psi$ are spelled out
in Kunen's.

>>>>

\renewcommand{\PP}{\mathbb{P}}
\renewcommand{\app}{\mathrm{App}}
\newcommand{\formula}{\ensuremath{\mathtt{formula}}}
\newcommand{\tyi}{\mathtt{i}}
\newcommand{\tyo}{\mathtt{o}}
\newcommand{\forceisa}{\mathtt{forces}}
\newcommand{\equ}{\mathbf{e}}
\newcommand{\bel}{\mathbf{b}}
\newcommand{\atr}{\mathit{atr}}
\newcommand{\reee}{\mathrel{\mathtt{eee}}}
\DeclareMathOperator{\val}{val}

@@

\subsection{Forcing for atomic formulas}
The base case of the definition of the forcing relation, that is, for
atomic formulas, is (in)famously the most complicated one. 
The definition adopted by Kunen~\cite[p.~257]{kunen2011set} is the
following, with two mutually recursive clauses:
\begin{multline}\label{eq:def-forcing-equality}
  \forceisa(\quine{t_1= t_2},\PP,\leq,\1,p) \defi 
  \forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :\\
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q)\lsii 
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q),
\end{multline}
\begin{multline}\label{eq:def-forcing-membership}
  \forceisa(\quine{t_1\in t_2},\PP,\leq,\1,p) \defi  \\
  \text{``}\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  \forceisa(\quine{t_1 =s},\PP,\leq,\1,q))\} \\
  \text{is dense below $p$.''}
\end{multline}
Other approaches, like the one in Neeman~\cite{neeman-course}, prefer
to have a single, more complicated, clause for $t_1=t_2$ using
well-founded recursion on $t_1$
and $t_2$ and then define the case $t_1\in t_2$ explicitly.

In any case, these definitions depend on the theorem of well-founded
recursion,
so the formula $\forceisa(\quine\phi,\PP,\leq,\1,p)$ for atomic $\phi$
must be obtained from the proof of that theorem. Below, a defined
relation $R$ is \emph{set-like} if $R^{-1}\{a\}$ is a set for every $a$.
%
\begin{theorem}[{\cite[I.9.11]{kunen2011set}}]
  Assume that $R$ is a
  set-like well-founded defined relation on a class $A$, assume that 
  $\forall  x,s\,\exists!y\,\phi(x,s,y)$ and write $G(x,s)$ for the
  only $y$ such that $\phi(x,s,y)$. Then there exists another formula 
   $\psi$ such that $\forall x\,\exists!y\,\psi(x,y)$ and if we write
  $F(x)$ for the only $y$ satisfying $\psi(x,y)$, we have
  \[
  \forall a\in A: F(a) = G(a,F\restr R^{-1}\{a\}).
  \]
\end{theorem}
%
Observe that since $F$ is a definable function, its restriction to the
set $R^{-1}\{a\}$ must also be a set by the Replacement Axiom.

There are several steps to apply this theorem to justify the
definition of forcing for atomic formulas:
\begin{enumerate}
\item As mentioned before, $F$ will be the characteristic function of
  the predicate $\forceisa$.
  %
\item The class $A$ and the relation $R$ must be
  defined.
  %
\item Write down explicitly $\psi$ as an element of
  $\formula$. 
\end{enumerate}
%
Luckily, the formula $\psi$ is spelled out in Kunen's.

Since the  function $F$ will essentially have 
the same  arguments  of $\forceisa$, but we will deconstruct the
formula and pack everything as a tuple. Namely, we will have
\begin{align}
  F(\lb\equ,t_1,t_2,\PP,\leq,\1,p\rb) =1 &\iff  
  \forceisa(\quine{t_1=t_2},\PP,\leq,\1,p) \label{eq:forcing-F-equ}\\
  F(\lb\bel,s,t_1,\PP,\leq,\1,p\rb) =1 &\iff  
  \forceisa(\quine{s\in t_1},\PP,\leq,\1,q). \label{eq:forcing-F-bel}
\end{align}
Here, $\equ$ and $\bel$ are arbitrary sets that encode which type of
formula we are considering. We can choose our $A$ to be
$\{\equ,\bel\}\times V^6$ (i.e., all 7-tuples whose first coordinate
are  $\equ$ or $\bel$).

Rewriting the clauses (\ref{eq:def-forcing-equality}) and
(\ref{eq:def-forcing-membership}), we might obtain a first
approximation to $G$. Notice that its domain is $A\times V$.

$G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =1$ if one of the following hold:
\begin{itemize}
\item $\atr=\equ$ and   $\forall s\in\dom(t_1)\cup\dom(t_2)\;\forall q\leq p :
  f(\lb\bel,s,t_1,\PP,\leq,\1,q\rb) =1  \lsii
  f(\lb\bel,s,t_2,\PP,\leq,\1,q\rb) =1$; or
\item $\atr=\bel$ and $\{q\leq p : \exists \lb s,r\rb\in t_2 (q\leq r \y
  f(\lb\equ,t_1,s,\PP,\leq,\1,q\rb) =1)\}$ is dense below $p$.
\end{itemize}
In any other case, $G(\lb\atr,t_1,t_2,\PP,\leq,\1,p\rb,f) =0$. In the
above, $f(v)=1$ should be interpreted as $\lb v,1\rb\in f$ (i.e., $f$
must be defined for the argument at hand). From this scheme an element
of $\formula$ must be produced.

The definition of the relation $R$ is straightforward and follows the
pattern of \cite[p.~257]{kunen2011set}. For instance, we have that 
\[
\lb\bel,s_1,t_1,\PP,\leq,\1,q\rb\mathrel{R}
\lb\equ,s_2,t_2,\PP,\leq,\1,p\rb
\]
if and only if 
\begin{itemize}
\item $s_1 \reee  s_2$ or $s_1\reee t_2$; and
\item $t_1 = s_2$ or $t_1 = t_2$.
\end{itemize}
%
\begin{framed}
  It seems necessary to prove that the relation $\reee$ is
  well-founded independently of universe where one is working. We only
  have a proof that ${\reee}(M)$ is well-founded for each $M$.
\end{framed}

With these elements we might reproduce the formula $\psi$ that defines
$F$ as in \cite[p.~49]{kunen2011set}. Let  $\app(d,h)$  be the formula
that states:
\begin{enumerate}
\item \label{item:dom-R-cerrado} $d$ is ``downwards $R$-closed'':
  $\forall y \in d, 
  R^{-1}\{y\}\subseteq d$;
\item \label{item:4} $h$ is a function with $\dom h =d \subseteq A$; and
\item \label{item:3} for all $y\in d$, $h(y) = G(y,h\restr(R^{-1}\{y\}))$.
\end{enumerate}
Then the formula defining $F$ is the following:
\[
(x\notin A \y y=0) \o \bigl(x\in A \y \exists d,h\, (\app(d,h)\y
x\in d \y h(x)=y)\bigr).
\]
The well-foundedness of $R$ is used to show that this formula is indeed
functional, and that it satisfies the required recursive equation
involving $G$. Once this is all translated to an element of \formula,
we'll be able to write the base case of the recursive definition of
$\forceisa$ using Equations~(\ref{eq:forcing-F-equ})
and~(\ref{eq:forcing-F-bel}).
